Deciding Bitvector Termination with Program Synthesis

### 2014-07-04 papers/termination/paper.tex
Kalashnikov soundness & completeness

\noindent {\bf Composing the environment with a loop.}
While proving that loop $L$ terminates, we create constraints on the environment in the
form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
must add constraints saying that $I$ overapproximates the states reachable via $L''$.
The details of how to create these constraints are standard, so
rather than presenting the algorithm for generating the constraints, we direct the
reader's attention to Figure~\ref{fig:environment-model}.  This figure contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
Note that despite its complexity, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

@@

\section{Soundness, Completeness and Complexity}
\begin{theorem}
\label{thm:sound}
 {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
 analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
 the program under analysis has at least one infinite execution.
\end{theorem}

@@

\begin{figure*}
\centering
\small
\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|r|}
\hline
    & LOC & Terminates? & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Res & Time (s) & Iterations\\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}

>>>>

\noindent {\bf Composing the environment with a loop.}
While proving that loop $L$ terminates, we create constraints on the environment in the
form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
must add constraints saying that $I$ overapproximates the states reachable via $L''$.
The details of how to create these constraints are standard, so
rather than presenting the algorithm for generating the constraints, we direct the
reader's attention to Figure~\ref{fig:environment-model}.  This figure contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

@@

\section{Soundness, Completeness and Complexity}
\begin{theorem}
 \label{thm:synth-sound}
 Algorithm~\ref{alg:abstract-refinement-code} is sound -- if it terminates with witness $P$ then
 $P \models \sigma$.
\end{theorem}

\begin{proof}
 The procedure {\sc RefinementLoop} terminates only if {\sc Verif} returns ``valid''.  In that
 case, $\exists x . \lnot \sigma(x, P)$ is unsatisfiable and so $\forall x . \sigma(x, P)$ holds.
\end{proof}

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable,
 algorithm~\ref{alg:abstract-refinement-code} is semi-complete -- if a program $P \models \sigma$
 exists then algorithm~\ref{alg:abstract-refinement-code} will terminate.  However if no program
 satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
 is finite, the algorithm is guaranteed to terminate.
\end{theorem}

\begin{proof}
 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked.  Since the first order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:abstract-refinement-code}
 enumerates this set, so it is semi-complete.

 If the domain $X$ is finite then the loop in procedure {\sc RefinementLoop} can only
 iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
 added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates (which it does
 by assumption of decidability), the whole algorithm terminates.
\end{proof}

\begin{theorem}
\label{thm:sound}
 {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
 analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
 the program under analysis has at least one infinite execution.
\end{theorem}

@@

\begin{figure*}
\centering
\footnotesize
\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|r|}
\hline
    & LOC & Terminates? & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Res & Time (s) & Iterations\\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}
